---
title: "Machine Learning Assignment"
author: "Vishnu Chevli"
---

- - - 

### Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 

This assignment deals with machine learning algorithm generation, which predict type of activity based on predictors.

```{r, echo=FALSE, results ='hide'}
library(knitr)
library(caret)
opts_chunk$set(eacho = TRUE, warning = FALSE, cache = TRUE)
```

- - -

### Data Processing

Let's start with data loading. We will also create a validation set with 80-20%

```{r}
# Loading training and testing file
training = read.csv('pml-training.csv', header= TRUE, stringsAsFactors = FALSE)
testing = read.csv('pml-testing.csv', header= TRUE, stringsAsFactors = FALSE)

# Convert outcome variable as factor 
training$classe = as.factor(training$classe)

# Setting seed for reproducibility
set.seed(123)

# From training set lets creat validation set with 80-20 ratio
validIdx = createDataPartition(y = training$classe, p = 0.8, list = FALSE)
validationSet = training[-validIdx,]
training = training[validIdx,]

rm(validIdx)             #   Removing variable
```

Let's find out required predictors from all predictors. Following code set will do cleaning as mentioned below
 1. Removing character predictors having blank values
 2. Finding numeric predictors having majority NAs
 3. Check for near zero variable checking

```{r}
# Majority of character variables are blank
ctype = data.frame(Variable = colnames(training), 
                   ctype = sapply(training, class, USE.NAMES = FALSE))
# Removing all charactor predicators except classes
training = training[,setdiff(colnames(training),
                          ctype[ ctype$ctype == 'character',]$Variable[-37])]

# Finding variable with maximum NA
resTable = sapply(training, function(x){sum(is.na(x))})

# There are variable with more than 90% of NAs removing them
training = training[,setdiff(colnames(training),
                             names(resTable[resTable >= (nrow(training)*0.9)]))]

# X, user_name and time won't be useful so removing those values from predicator
training = training[,setdiff(colnames(training), 
                             c('X', 'raw_timestamp_part_1', 
                               'raw_timestamp_part_2'))]

# Near zero checking, as results are negetive so avoiding 
nearZeroTrain <- nearZeroVar(training[,-54], saveMetrics = TRUE)

# Removing unnecessary variables
rm(resTable, ctype, nearZeroTrain)
```

We have already discarded more than 100 predictors. Now we have only 53 numeric predictors, but still this number is very big. Lets do principle component analysis and keep only those component with only 80% variance (~90% standard deviation)

```{r}
# Setting seed for reproducibility
set.seed(123)

# Lets create principal component with 80% threshold
preCompAna = preProcess(training[,-54], method = 'pca', thresh = 0.8)

# Find out actual values using principal component
trainPC <- predict(preCompAna, training[,-54])
```

- - -

### Model Fitting

We have now 13 final predictors. We will fit random forest using train function for training set.

```{r}
# Setting seed for reproducibility
set.seed(123)

# Let's fit our model using 
fit <- train(training$classe~., data = trainPC, method = 'rf')

# Let's check confusion matrix with training values
confusionMatrix(training$classe,predict(fit,trainPC))
```

Now let's check our model fit with validation set.

```{r}
# Creating components for validationSet using training set
validPC <- predict(preCompAna, validationSet[,colnames(training)[-54]])

# Comparing prediction with actual values in validationSet
confusionMatrix(validationSet$classe, predict(fit, validPC))
```

Let's predict values for test set.

```{r}
# Creating components for testing set using training set components
testPC <- predict(preCompAna, testing[,colnames(training)[-54]])

# Printing predicted values
predict(fit, testPC)
```